Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2019-01-15 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

We are grateful to the reviewers for their helpful comments. Here we submit a very substantial revision implementing almost all of them.

We realize that this paper is being resubmitted well beyond the typical deadline for responses. We apologize for that, however we hope the reviewers understand that in the interim the first author has gotten an industry job, we have added an author, and we have rewritten a sizeable fraction of the proofs.

@@

Outline of major revisions:
1. We have generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal. The previous version presented an argument that the ML branch lengths would have this property, but this argument did not hold up to further scrutiny.
2. We have expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation.
3. We have changed discussion and plots to involve branch lengths and not fidelities.

Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach. Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done. As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al. We provide some numerical results to confirm our derivations and to provide a comparison.

This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section. Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice. An investigation of bigger trees and more complex models deserves another paper. We have added clarification of these points to the discussion: "All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners."

@@

In the revised manuscript, all region bounds are analytical as we have determined a closed form solution for the joint inference problem in this case.

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

We have added the line:
"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."

We believe it's important not to mischaracterize the suggested use of joint inference, though we would like to emphasize to the reader that problems are apparent in large regions of the parameter space. We think the current discussion treats this optimization procedure as fairly as possible, although we would gladly accept any further specific suggestions.

@@

> Comments to the Author
> Review of Shaw, D. and Matsen, F.  "Joint maximum-likelihood of
> phylogenies and ancestral states is not consistent"

> Shaw and Matsen demonstrate that the approach of jointly estimating
> ancestral states and tree topologies (or branch lengths on a fixed
> topology) is not statistically consistent.  I believe that their
> results are correct, and that their result is important to
> disseminate.  The paper is clearly written, given the complexity of
> the material, and I have no reservation about recommending publication
> virtually as is.

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

Thank you for your assessment. We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and hope that reviewers will be able to read the revision more fully.

### 2019-03-19 versions/response.v2.txt

> The reviewers have universally viewed the revision as much improved.
> That said, 3/4 maintain that further changes are necessary prior to
> publication. It is our view that the remaining criticisms are "minor",
> and should you address them thoroughly and convincingly, we expect
> that final acceptance of the revision can be done editorially.

XXX

> Reviewers' comments:
> Reviewer: 1

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

TODO: add discussion of bootstrap

> Getting complete results about topological consistency is feasible.
> The arguments that led to the closed form solution for the limiting
> edge-length estimates for tau_1 can be repeated similarly for tau2.
> This would allow, for any given x* & y*, comparison of the limiting
> lnl per site for tau1 vs tau2. Whichever topology has the larger lnl
> is the long-run estimated topology. In cases where hat w=1 for both
> topologies, conclusions are different but definite. Thus it seems a
> relatively straightforward exercise given the work the authors have
> already done (I hope I am not missing a subtle difficulty) to have a
> computer determine for a grid of (x*,y*) whether (a) hat w=1 exactly
> for both topologies, (b) the difference in limiting lnl over
> topologies is either positive or negative or (c) hat w neq 1 for at
> least one of the topologies and the difference in limiting lnl is 0 up
> to machine precision. I would emphasize that the cases (a) and (b)
> constitute rigorous proof. The cases (c) likely are rare and
> correspond to boundary cases, so the boundary may have some slight
> uncertainty. There may be some aesthetic appeal to getting analytical
> results rather than using a computer but being able to draw clear
> conclusions about strict topological inconsistency is important.

DISCUSS: we have no evidence of topological inconsistency

> The language used to refer to the approach and references to the
> literature are awkward. The references to profile likelihood work
> suggest that there is a theoretical basis for the approach (that paper
> suggests consistency) but the settings considered in such work are
> quite different, I believe. Integrated likelihood usually refers to
> something different too. Finally, as Reviewer 3 makes clear the only
> current implementation of the approach was intended as an
> approximation to ML rather than as ML. The central issue is whether
> treating unobserved data (ancestral states here) as parameters
> sometimes works; the EM algorithm were developed with the idea that
> this is not what should be done. In any case, references directly tied
> to that issue would be more relevant. Additional suggestions include:
> 1) Make a clear statement in the intro about why the usual likelihood
> is considered the right likelihood: it is the probability of the
> *observed* data. 2) The point that large numbers of parameters need
> not defeat likelihood methods is valid but it should be made clear
> when referencing such work that the setting is different.

> -l21: Some rewording is needed. The joint method will, as n gets
> large, continue to estimate a well-resolved topology with positive
> probability; albeit with small middle edge-length.

> -l131: A reader might be able to avoid the appendix (which is
> difficult to wade into) if an example zeta in a four taxon case was
> given as illustration.

> -l150: I initially found this paragraph confusing. Table S1 is a
> result that you are using as intuition, so you can't be giving
> intuition for the entire result (otherwise it gets circular). I think
> you are trying to give intuition for the `everywhere except a set of
> measure' portion of the result rather than entire result. Perhaps `the
> proof that hat t neq t* except on a set of measure 0 can be seen
> intuitively...'.

> -l154 and (should check elsewhere): `the limiting value of the
> estimator' rather than `the estimator' is what is being referred to
> here.

> -l186: which `likelihood terms' is not that clear to me: `each site
> results in multiple factors, only one of which depends on w with
> dependence being through (1+w) or (1-w) depending ...'

> -l189: `depends on the values of other branch parameters' or some
> other rewording. I don't think it is that they are unknown. It is that
> they need optimization, so you can't optimize w independently of them.

> -l201, Figs 2 and 3: I don't understand why direct numerical
> optimization was needed. You have the exact expressions for hat w in
> Fig S2. So you could have plotted them as Fig 2. A similar comment
> applies to Fig 3. Are you trying to convince the reader that your
> proofs have no errors? That seems more an issue for supp material.

> -l252-259: I don't think this is an example of profile likelihood as
> considered in Murphy and Van der Vaart.

> -l264: Do you mean that `consistency can hold even when the numbers of
> parameters tend to infinity'? The statement `not a necessary
> condition' just means that inconsistency can arise in different ways
> (eg. lack of identifiability).

> Reviewer: 2

> Comments to the Author
> Shaw et al provided a revised version of the manuscript that I
> reviewed about a year ago showing that joint estimation of ancestral
> states and branch length is in general not consistent. The authors now
> map out the regions in which topologies are incorrectly estimated and
> include a plot that shows the bias of the internal branch length for
> different tree configurations. I think this is a useful piece of work.
> But some parts are completely unsurprising, while some empirical
> results seem inaccurate or misleading.

> The most important results/claims by Shaw et al are:

> 1) "jointly estimating phylogenetic parameters along with ancestral
> states is not consistent in general"... "only correct on a set of
> measure 0".
> It is clearly true that branch length will be biased. But the fact
> that joint estimation will not recover the true branch lengths is not
> surprising but perfectly expected. Joint estimation ignores subleading
> configurations of internal states. Since the subleading terms
> typically require more state changes, branch length estimates ignoring
> them are biased. In a similar way a Taylor expansion is almost nowhere
> "correct", but still a useful and insightful approximation (a better
> analogy is ignoring looped diagram in a diagrammatic expansion.).

> 2) In some parameter regions the internal branch of the four taxon
> tree collapses to 0.
> Again, this is true and identifying regions in parameter space where
> the internal branch collapses to zero is the most interesting part of
> the manuscript in my opinion. Unfortunately, the manuscript doesn't
> give much insight as to why the length of the internal branch shrinks
> to zero in some regions of parameter space. There is, however, a
> rather intuitive explanation for this. In the inv Fels topology, the
> internal branch collapses to 0 if
>
> p_x^2 > c p_y
>
> (the prefactor c depends on specifics like alphabet size etc). And
> this has an intuitive interpretation: if p_x^2 > p_y, it is more
> likely to observe a mutation twice on two branches of length p_x than
> once on a branch of length p_y. As a result, the joint inference will
> push p_w to zero and assign all state changes on the terminal branches
> of length p_x. But for short branches, this corresponding sector of
> parameter space is a small (tending to zero for a fixed p_x/p_y with
> p_x \to 0).
>
> My empirical tests of the approximate joint inference scheme suggest
> this region is much smaller than the one identified by Shaw et al in
> Fig 3.

> 3) "[The bias is] of the same order as the branch lengths themselves."
> This is incorrect or misleading, depending on how you read it. In
> typically cases, the error made by this procedure should be quadratic
> in branch length and hence of higher order than the branch length
> itself. All empirical test we have done via numerically optimizing
> branch lengths suggests this is true for typical trees and that the
> errors made are small provided ALL adjacent branch lengths are short.
> That said, relative error of one branch can be large if adjacant
> branches are long such as the case above. In this case the error is
> quadratic in the length of the longer adjacent branches rather than
> the branch itself.

> Reviewer: 3

> Comments to the Author
> This a resubmission of an earlier version of the paper, for which I
> was reviewer #4.  The other reviewers of the previous submission did
> an excellent job in pointing out minor problems and making suggestions
> for improvement.  It seems that Shaw et al. have addressed all of
> these suggestions in exemplary fashion, and  think the result is a
> much stronger paper.  I especially appreciate the authors' efforts to
> make the mathematical derivations and proofs more accessible to
> general readers (like myself).

> I often write long reviews but given the careful attention to detail
> in responding to the previous reviews of an already good paper, I have
> nothing further to add.

> Reviewer: 4

> Comments to the Author
> The authors have made some substantial changes addressing the concerns
> of the last round and I think that has improved the clarity and in
> some instances also strengthened the results.
> Remaining comments are mostly about the notation which I still find a
> bit hard to follow.
> Why are site and ancestral node splits carrying this distinction of
> whether the m'th taxon (or node) has a 1?

> Perhaps it's a latexdiff leftover but eq 6 in line 151 seems to have a
> line with just a ; \xi )

> Line 593 p 25 I don't understand "the mapping \psi operates by
> returning the tips labeled as 1 in a site pattern to obtain a site
> split in P(Y) if the set of tips labeled 1 is not in P(Y)" and this
> mapping seems fundamental. Perhaps you mean to refer to the m'th taxon
> (or node)..?

> The symmetric model is assumed strongly in the first part of the
> appendix, with the factor of 2; my guess is that the results hold in
> asymmetric models but the notation is harder, is that correct?

> What's q in equation 7 in the appendix (line 655) ?

> More substantitvely, the limit in (8) seems not established because as
> n -> \infty so would q and n_j, so these can't be treated as fixed .

> page 9 - at this point in the paper, unless the reader has already
> read the appendix they don't know what the estimator means or what the
> p_ytilde j are - so this text doesn't make sense at this point.

> page 30 line 667-8 what are S and Y and what makes a path? I gather
> from the math that you mean paths in the tree, since the sum is over
> ancestral states and there's a product of probabilities of changes on
> branches but (perhaps my lack of background) the form of (9) isn't
> obvious I think because I am not exactly clear on what the notation
> means.

### 2019-03-19 versions/joint_inf.v2.tex
Adding response template and old version of ms

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we show empirically that branch lengths are biased in this setting even for short branches, with bias of the same order as the branch lengths themselves.
\end{abstract}

@@

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
Although the regions seem coarse, $\hat{w}$ has a closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
%EM I don't feel like this "range" is a very strong statement. What if we just repeat that the bias is of the same order as the branch lengths and give an example?
%das: that works, though I'm not too clear on what kind of example you're talking about. Maybe Fig. 3 is enough?
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to provably consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
We show empirically that branch length estimates are biased even for short branches, with bias of the same order as the branch lengths themselves.
\end{abstract}

@@

This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
